preciseTAD: A machine learning framework for precise TAD boundary prediction at base pair resolution

### 2020-07-31 Manuscript/manuscript.Rmd
updated results

---
title: "preciseTAD: A machine learning framework for precise TAD boundary prediction at base pair resolution"
csl: styles.ref/genomebiology.csl
output:
  word_document:
    reference_docx: styles.doc/NIH_grant_style.docx
  pdf_document: default
bibliography: References/references.bib
---

@@

Within this study we used both TAD boundary and loop boundary data derived from the publicly available tools ARROWHEAD [@durand2016juicebox] and Peakachu [@salameh2019supervised] respectively. ARROWHEAD TAD boundaries were called from primary Hi-C data from Geo GSE63525 [@rao20143d] for the GM12878 and K562 cell lines (Supplementary Table 1). The contact matrices were constructed from all read pairs that uniquely aligned to the genome (MAPQ>0). The default tools for ARROWHEAD were used when extracting the autosomal genomic coordinates (GRCh37/hg19 human genome assembly) of TAD boundaries at 5 kb, 10 kb, 25 kb, 50 kb, and 100 kb resolutions (Additional File 1). Peakachu chromatin loop boundaries called at 10 kb for the GM12878 and K562 cell lines were downloaded from the Yue lab web site (Supplementary Table 1). Unique boundaries were considered as the midpoints within the coordinate of each chromatin loop. Chromosome 9 was excluded from all downstream analyses due to sparsity of contact matrices at 5 kb and 10 kb resolutions for the K562 cell line. Cell-line-specific genomic annotations (BroadHMM chromatin segmentations (BroadHMM), histone modifications (HM), and transcription factor binding sites (TFBS)) were obtained from the UCSC Genome Browser Database [@Tyner:2017aa]  (GRCh37/hg19 human genome assembly, Supplementary Table 2).

@@

Next, we assessed the biological significance of our predicted boundaries with respect to the enrichment profiles of several known molecular drivers of 3D chromatin architecture including CTCF, RAD21, SMC3, and ZNF143 ChIP-Seq peaks. To do so, we used the computeMatrix [??? Again, I expected italics, but decide and correct;!!!talk about in meeting] and plotHeatmap tools offered by deepTools version 2.0 [@ramirez2016deeptools2] via integration with Galaxy version 3.3.2.0.0 [@afgan2018galaxy]. Additionally, we compared the median $log_2$ genomic distances between TAD boundaries and the same top predictive ChIP-seq annotations using Wilcoxon Rank-Sum test. Furthermore, we compared the overlap between predicted and called boundaries in GM12878 and K562 cell lines. Boundaries were first flanked by resolution, *r*, and overlaps were visualized using Venn diagrams from the `Vennerable` R package (version 3.1.0). Overlaps were further quantified using Jaccard index defined as

@@

We developed a machine learning framework<!--, refered to as *TADRF*,--> aimed at determining the optimal set of data level characteristics necessary for topologically associated domain (TAD) boundary prediction. We chose the random forest (RF) machine learning algorithm as our binary classification tool. The reason for choosing RF was two-fold: (1) In order to devise a tunable prediction rule in a supervised learning framework that is both robust to overfitting and able to handle multiple correlated predictors, and (2) to allow for an interpretable ranking of predictors with respect to their ability to predict boundary regions [@boulesteix2012overview]. We then developed *preciseTAD*, an algorithm which leverages an optimally built RF to precisely predict the genomic coordinates of both TAD and chromatin loop boundaries at base pair resolution.

@@

Boundary regions were defined as regions containing a called boundary ($Y = 1$), while non-boundary regions were defined as genomic bins that did not contain a called boundary ($Y = 0$) (Figure 1A, see Methods). The total number of TADs, TAD boundaries, and the number of genomic bins expectedly decreased with the decreased resolution of Hi-C data (Table 1, Supplementary Table 2 [??? I can see numbers for K562 (?) only? There's a discrepancy with legend.!!!fixed, need to added spectralTAD]). The number of non-boundary regions highly outnumbered boundary regions, with results consistent for loop boundaries as well (Supplementary Table 3). Such disproportional presence of examples in one class is known as a "class imbalance" problem that negatively affects predictive modeling [@prati2011survey; @wei2013role]. To address class imbalance, we evaluated the effect of three re-sampling techniques. *Random over-sampling* (ROS) was defined as sampling with replacement from the minority class (boundary regions). *Random under-sampling* (RUS) was defined as sampling with replacement from the majority cluss (non-boundary regions). Lastly, we tested *Synthetic minority over-sampling technique* (SMOTE), which is a combination of both random over- and under-sampling to create balanced classes [@chawla2002smote] (see Methods).

A total of 77 [??? 77 in the table?;!!!fixed] cell-line-specific functional genomic annotations were used to build the predictor space. These included BroadHMM chromatin segmentations, histone modifications (HM), and transcription factor binding sites (TFBS) (Supplementary Table 4). Four feature engineering procedures were developed to quantify the association between genomic annotations and bins (Figure 1B). These include signal strength association (Signal), direct (OC), proportional (OP), and spatial (Distance) relationships (see Methods).

![**Resolution-specific data construction and feature engineering used for downstream random forest modeling.** (A) The linear genome was binned into non-overlapping resolution-specific intervals (see Methods). The response vector **Y** was defined as 1/0 if a genomic bin overlapped/did not overlap with a TAD boundary, respectively. (B) Four types of associations between bins and genomic annotations (green shapes) were considered to build the predictor space including _Average Peak Signal_ (Signal), _Overlap Counts_ (OC), _Overlap Percent_ (OP), and ($log_{2}$) _Distance_.](Figures/figure1_model_construction_and_predictor_types.png)

@@

![**A diagram of the machine learning framework.** Step 1 defines the chromosome-specific predictor-response matrix **A** of $N$ genomic bins (rows), $p$ genomic annotations, and one response vector $Y$ ($p+1$ columns). Resolution-specific data (5 kb, 10 kb, 25 kb, 50 kb, or 100 kb) and feature engineering procedure (overlap count (OC), overlap percent (OP), $log_{2}$ distance (Distance)) for the $n-1$ autosomal chromosomes. Step 2 is to reserve the predictor-response matrix for the holdout chromosome $i$ as the test data. Step 3 is to apply a re-sampling technique to the training set (no re-sampling, random over-sampling, random under-sampling, or SMOTE re-sampling). Step 4 is to train the random forest model and perform 3-fold cross-validation to tune for the *mtry* parameter. Finally, step 5 is to validate the model on the separate test data and evaluate model performance using balanced accuracy (BA).](Figures/figure2_ensemble_framework.png) [??? Define all indices, what is N, p+1, i?;!!!defined in methods, define here again?]

@@

| Resolution | Total number of called TADs | Total number of unique TAD boundaries | Total number of genomic bins | Class imbalance |
|:----------:|:---------------------------:|:-------------------------------------:|:----------------------------:|:---------------:|
|5 kb        |8052                         | 15468                                 | 535363                       | 0.03            |
|10 kb	     |7676 	                       | 14253                                 | 267682                       | 0.05            |
|25 kb	     |4670                         | 8363                                  | 107073                       | 0.08            |
|50 kb	     |2349	                       | 4224                                  | 53537                        | 0.08            |
|100 kb	     |1031                         | 1883                                  | 26768                        | 0.07            |
Table: Domain data and class imbalance summaries across resolutions for ARROWHEAD on the GM12878 cell line.

@@

![**???[needs updating with new framework, takes longest, do last - ??? Is this current?;talk about in meeting] Distance-type predictors combined with random under-sampling yielded optimal TAD boundary region predictive performance for GM12879 ARROWHEAD TAD boundaries.** Balanced accuracy is compared across resolution, within each predictor-type: (A) Signal, (B) OC, (C) OP, and (D) Distance, and across re-sampling techniques: no re-sampling (None; red), random over-sampling (ROS; green), random under-sampling (RUS; blue), and synthetic minority over-sampling (SMOTE; purple).](Figures/figure3_performances.png)

@@

![**Binding of SMC3, Rad21, CTCF, and Znf143 transcription factors accurately predicts TAD and loop boundaries in GM12878 cell line.** (A) Barplots comparing performances of TAD (ARROWHEAD) and loop (Peakachu) boundary prediction models using histone modifications (HM), cis-regulated elements (BroadHMM), transcription factor binding sites (TFBS), in addition to a model containing all three classes (ALL). (B) Recursive feature elimination (RFE) analysis used to select the optimal number of predictors. Error bars represent standard deviation from the cross-validated accuracy across each chromosome. (C) Clustered heatmap of the predictive importance for the union of the top 8 chromosome-specific most predictive TFBSs.](Figures/figure4_annots_specifc_rfe_importances_arrowhead_peakachu.png)

@@

![**preciseTAD boundaries are enriched in CTCF sites.** (A) The location of preciseTAD (green) vs. ARROWHEAD (blue) boundaries on GM12878 (chr 1: 70600000-71530000). The black line represents the probability of each base being a TAD boundary. The green vertical lines represent the *preciseTAD boundary points* (PTBP). A zoomed-in portion of the genome shows a yellow-highlighted cluster of bases with high probability of being a boundary, the *preciseTAD boundary region (PTBR). (B) Heatmaps comparing the enrichment of CTCF signal around preciseTAD and ARROWHEAD TAD boundaries, and (C) around preciseTAD and Peakachu loop boundaries. (D) $Log_{2}$ genomic distances from boundaries to nearest CTCF sites. (E) Pairwise Jaccard indices shows better overlap among *preciseTAD*-detected boundaries across resolutions (upper triangular) as compared to ARROWHEAD (lower triangular).](Figures/figure5_enrichment_and_log2dist_ctcf.png) [??? Panel E - do we have it for Peakachu? - stretched, can be put in figure. Can you color triangulars blue/green for Arrowhead/preciseTAD, green for preciseTAD?;!!!can't be done for peakachu because we only have 1 resolution (10kb)]

@@

![**preciseTAD boundaries are more conserved between the GM12878 and K562 cell lines.** (A) Venn diagrams illustrating the overlap between GM12878- and K562-specific predicted and called boundaries. (B) Venn diagrams illustrating the agreement between boundaries predicted using ARROWHEAD- and Peakachu-trained models.](Figures/figure6_venn_diagrams.png) 
[??? What is the difference between leftmost Venn diagrams on panel A?] [??? What are the numbers?;!!!talk in meeting; add indication of what ground truth is for preciseTAD]

@@

**Supplementary Table 2.** Domain boundary and class imbalance summaries from ARROWHEAD for the K562 cell line. [???add spectralTAD]

**Supplementary Table 3.** Loop boundary and class imbalance summaries from Peakachu 10 kb resolution for both the GM12878 and K562 cell lines respectively.

**Supplementary Table 4.** A complete list of the genomic annotations used to build the predictor space for all downstream models. The GRCh37/hg19 human genome assembly was used. "Genomic Class" - broad category of genomic features, "Element" - names of genomic features, "Cell Type-Specific Source" - download URL specific to the cell line (not all annotations were provided by the same institutions).

@@

**Supplementary Figure 2. Distance-type predictors combined with random under-sampling yielded optimal TAD boundary region predictive performance for K562 ARROWHEAD TAD boundaries.** Balanced accuracy is compared across resolution, within each predictor-type: (A) Signal, (B) OC, (C) OP, and (D) Distance, and across re-sampling techniques: no re-sampling (None; red), random over-sampling (ROS; green), random under-sampling (RUS; blue), and synthetic minority over-sampling (SMOTE; purple).

**Supplementary Figure 3. Distance-type predictors combined with random under-sampling or SMOTE re-sampling yielded optimal predictive performance for A) GM12878 and B) K562 Peakachu loop boundaries.** Balanced accuracy is compared within each predictor-type: Signal, OC, OP, and Distance, and across across resampling technique: no re-sampling (None; red), random over-sampling (ROS; green), random under-sampling (RUS; blue), and synthetic minority over-sampling (SMOTE; purple).

**Supplementary Figure 4. Binding of SMC3, Rad21, CTCF, and Znf143 transcription factors accurately predicts TAD and loop boundaries in K562 cell line.** (A) Barplots comparing performances of TAD (ARROWHEAD) and loop (Peakachu) boundary prediction models using histone modifications (HM), cis-regulated elements (BroadHMM), transcription factor binding sites (TFBS), in addition to a model containing all three classes (ALL). (B) Recursive feature elimination (RFE) analysis used to select the optimal number of predictors. Error bars represent standard deviation from the cross-validated accuracy across each chromosome. (C) Clustered heatmap of the predictive importance for the union of the top 8 chromosome-specific most predictive TFBSs.

**Supplementary Figure 8. preciseTAD boundaries are more enriched for GM12878-specific SMC3, RAD21, and ZNF143 sites.** Enriched heatmaps comparing preciseTAD boundaries with both (A) ARROWHEAD and (B) Peakachu boundaries for the GM12878 cell line.

**Supplementary Figure 9.preciseTAD boundaries are more enriched for K562-specific CTCF, SMC3, RAD21, and ZNF143 sites.** Enriched heatmaps comparing preciseTAD boundaries with both (A) ARROWHEAD and (B) Peakachu boundaries for the K562 cell line.

**Supplementary Figure 10. preciseTAD boundaries are spatially closer to GM12878-specific SMC3, RAD21, and ZNF143 sites.** Boxplots comparing the $log_{2}$ genomic distances from each preciseTAD boundary (green) and called boundary (either ARROWHEAD or Peakachu; blue) to the nearest SMC3, RAD21, and ZNF143 sites on the linear genome for GM12878.

**Supplementary Figure 11. preciseTAD boundaries are spatially closer to K572-specific CTCF, SMC3, RAD21, and ZNF143 sites.** Boxplots comparing the $log_{2}$ genomic distances from each preciseTAD boundary (green) and called boundary (either ARROWHEAD or Peakachu; blue) to the nearest CTCF, SMC3, RAD21, and ZNF143 sites on the linear genome for K562.
[??? Combine Supplementary Figure 10 and 11. Make panel A and, underneath, panel B. Remove CTCF from Y-axis and put overhead labels above each plot pair]

**Supplementary Table 5.** Median chromosome-specific Jaccard indices (IQR) for ARROWHEAD- and Peakachu-called and *preciseTAD* predicted boundaries between the GM12878 and K562 cell lines [??? Where is cell line information?] [??? What is in parentheses?]. The p-value is from a nonparametric Wilcoxon-rank sum test [??? Absent]. Note, ARROWHEAD-called and predicted boundaries were flanked by 5 kb, while Peakachu-called and predicted boundaries were flanked by 10 kb prior to calculating Jaccard indices.

**Supplementary Table 6.** Additional performance metrics when implementing a random forest using ARROWHEAD ground truth TAD boundaries.

**Supplementary Table 7.** Additional performance metrics when implementing a random forest using Peakachu ground truth chromatin loop boundaries.

>>>>

---
title: "preciseTAD: A machine learning framework for precise TAD boundary prediction at base pair resolution"
csl: styles.ref/genomebiology.csl
output:
  pdf_document: 
  word_document:
    reference_docx: styles.doc/NIH_grant_style.docx
bibliography: References/references.bib
---

@@

Within this study we used both TAD boundary and loop boundary data derived from the publicly available tools ARROWHEAD [@durand2016juicebox] and Peakachu [@salameh2019supervised] respectively. ARROWHEAD TAD boundaries were called from primary Hi-C data from Geo GSE63525 [@rao20143d] for the GM12878 and K562 cell lines (Supplementary Table 1). The contact matrices were constructed from all read pairs that uniquely aligned to the genome (MAPQ>0). The default parameters for ARROWHEAD were used when extracting the autosomal genomic coordinates (GRCh37/hg19 human genome assembly) of TAD boundaries at 5 kb, 10 kb, 25 kb, 50 kb, and 100 kb resolutions (Additional File 1). Peakachu chromatin loop boundaries called at 10 kb for the GM12878 and K562 cell lines were downloaded from the Yue lab web site <!--(Supplementary Table 1)-->. Unique boundaries were considered as the midpoints within the coordinate of each chromatin loop. Chromosome 9 was excluded from all downstream analyses due to sparsity of contact matrices at 5 kb and 10 kb resolutions for the K562 cell line. Cell-line-specific genomic annotations (BroadHMM chromatin segmentations (BroadHMM), histone modifications (HM), and transcription factor binding sites (TFBS)) were obtained from the UCSC Genome Browser Database [@Tyner:2017aa]  (GRCh37/hg19 human genome assembly<!--, Supplementary Table 2-->).

@@

Next, we assessed the biological significance of our predicted boundaries with respect to the enrichment profiles of several known molecular drivers of 3D chromatin architecture including CTCF, RAD21, SMC3, and ZNF143 ChIP-Seq peaks. To do so, we used the computeMatrix [??? Again, I expected italics, but decide and correct;!!!talk about in meeting] and plotHeatmap tools offered by deepTools version 2.0 [@ramirez2016deeptools] via integration with Galaxy version 3.3.2.0.0 [@afgan2018galaxy]. Additionally, we compared the median $log_2$ genomic distances between TAD boundaries and the same top predictive ChIP-seq annotations using Wilcoxon Rank-Sum test. Furthermore, we compared the overlap between predicted and called boundaries in GM12878 and K562 cell lines. Boundaries were first flanked by resolution, *r*, and overlaps were visualized using Venn diagrams from the `Vennerable` R package (version 3.1.0). Overlaps were further quantified using Jaccard index defined as

@@

We developed a machine learning framework aimed at determining the optimal set of data level characteristics necessary for topologically associated domain (TAD) boundary prediction. We chose the random forest (RF) machine learning algorithm as our binary classification tool. The reason for choosing RF was two-fold: (1) In order to devise a tunable prediction rule in a supervised learning framework that is both robust to overfitting and able to handle multiple correlated predictors, and (2) to allow for an interpretable ranking of predictors with respect to their ability to predict boundary regions [@boulesteix2012overview]. We then developed *preciseTAD*, an algorithm which leverages an optimally built RF to precisely predict the genomic coordinates of both TAD and chromatin loop boundaries at base pair resolution.

@@

Boundary regions were defined as regions containing a called boundary ($Y = 1$), while non-boundary regions were defined as genomic bins that did not contain a called boundary ($Y = 0$) (Figure 1A, see Methods). The total number of TADs, TAD boundaries, and the number of genomic bins expectedly decreased with the decreased resolution of Hi-C data (Table 1, Supplementary Table 2 [??? I can see numbers for K562 (?) only? There's a discrepancy with legend.!!!fixed, need to added spectralTAD]). The number of non-boundary regions highly outnumbered boundary regions, with results consistent for loop boundaries as well. Such disproportional presence of examples in one class is known as a "class imbalance" problem that negatively affects predictive modeling [@prati2011survey; @wei2013role]. To address class imbalance, we evaluated the effect of three re-sampling techniques. *Random over-sampling* (ROS) was defined as sampling with replacement from the minority class (boundary regions). *Random under-sampling* (RUS) was defined as sampling with replacement from the majority cluss (non-boundary regions). Lastly, we tested *Synthetic minority over-sampling technique* (SMOTE), which is a combination of both random over- and under-sampling to create balanced classes [@chawla2002smote] (see Methods).

A total of 77 [??? 77 in the table?;!!!fixed] cell-line-specific functional genomic annotations were used to build the predictor space. These included BroadHMM chromatin segmentations, histone modifications (HM), and transcription factor binding sites (TFBS) (Supplementary Table 3). Four feature engineering procedures were developed to quantify the association between genomic annotations and bins (Figure 1B). These include signal strength association (Signal), direct (OC), proportional (OP), and spatial (Distance) relationships (see Methods).

![**Resolution-specific data construction and feature engineering used for downstream random forest modeling.** (A) The linear genome was binned into non-overlapping resolution-specific intervals (see Methods). The response vector **Y** was defined as 1/0 if a genomic bin overlapped/did not overlap with a TAD boundary, respectively. (B) Four types of associations between bins and genomic annotations (green shapes) were considered to build the predictor space including Average Peak Signal (Signal), Overlap Counts (OC), Overlap Percent (OP), and ($log_{2}$) Distance.](Figures/figure1.png)

@@

![**A diagram of the machine learning framework used to build TAD/loop boundary region prediction models.** Step 1 defines the data matrix A, composed of N genomic bins for each chromosome, except a holdout chromosome *i*, *p* genomic annotations, and the response vector **Y** (for *p*+1 columns). Resolution-specific data (5 kb, 10 kb, 25 kb, 50 kb, or 100 kb) and feature engineering procedure (overlap count (OC), overlap percent (OP), $log_{2}$ distance (Distance)) for the n-1 autosomal chromosomes. Step 2 is to reserve the predictor-response matrix for the holdout chromosome *i* as the test data. Step 3 is to apply a re-sampling technique to the training set (no re-sampling, random over-sampling, random under-sampling, or SMOTE re-sampling). Step 4 is to train the random forest model and perform 3-fold cross-validation to tune for the mtry parameter. Finally, step 5 is to validate the model on the separate test data composed of the binned data from the holdout chromosome *i* and to evaluate model performance using balanced accuracy (BA).](Figures/figure2.png)

@@

|          Tool         |     Resolution    |     Total number of called TADs/chromatin loops    |     Total number of unique TAD/chromatin loop boundaries    |     Total number of genomic bins    |     Class imbalance    |
|:---------------------:|:-----------------:|:--------------------------------------------------:|:-----------------------------------------------------------:|:-----------------------------------:|:----------------------:|
|        ARROWHEAD      |       5   kb      |                         8052                       |                             15468                           |                535363               |           0.03         |
|     ARROWHEAD         |       10   kb     |                         7676                       |                             14253                           |                267682               |           0.05         |
|     ARROWHEAD         |       25   kb     |                         4670                       |                             8363                            |                107073               |           0.08         |
|     ARROWHEAD         |       50   kb     |                         2349                       |                             4224                            |                 53537               |           0.08         |
|     ARROWHEAD         |      100   kb     |                         1031                       |                             1883                            |                 26768               |           0.07         |
|        PEAKACHU       |       10   kb     |                        16185                       |                          21421                              |                267682               |        0.14            |
Table: TAD/loop boundary data and class imbalance summaries across resolutions for ARROWHEAD and PEAKACHU on the GM12878 cell line. Class imbalance is calculated as the ratio of total number of unique boundaries to the total number of genomic bins.

@@

![*Distance-type predictors combined with random under-sampling at 5 kb resolution genomic bins yielded optimal TAD boundary region predictive performance for GM12878 ARROWHEAD TAD boundaries.** Averaged balanced accuracies are compared across resolution, within each predictor-type: (A) Signal, (B) OC, (C) OP, and (D) Distance, and across re-sampling techniques: no re-sampling (None; red), random over-sampling (ROS; green), random under-sampling (RUS; blue), and synthetic minority over-sampling (SMOTE; purple). Error bars indicate 1 standard deviation away from the mean performance across each holdout chromosome used for testing.](Figures/figure3.png)

@@

![**Binding of SMC3, RAD21, CTCF, and ZNF143 transcription factors accurately predicts TAD and loop boundaries in GM12878 cell line.** (A) Barplots comparing performances of TAD (ARROWHEAD) and loop (PEAKACHU) boundary prediction models using histone modifications (HM), cis-regulated elements (BroadHMM), transcription factor binding sites (TFBS), in addition to a model containing all three classes (ALL). (B) Recursive feature elimination (RFE) analysis used to select the optimal number of predictors. Error bars represent 1 standard deviation from the cross-validated accuracy across each holdout chromosome. (C) Clustered heatmap of the predictive importance for the union of the top 8 most predictive chromosome-specific TFBSs. The columns represent the holdout chromosome excluded from the training data. (D) Receiver operating characteristic (ROC) curves and subsequent area under the curves (AUCs) when training and testing on the same cell line versus across cell lines. The curves represent the average sensitivities and specificities across each holdout chromosome.](Figures/figure4.png)

@@

![**preciseTAD boundaries are enriched for known molecular drivers of 3D chromatin.**(A) The location of preciseTAD (green) vs. ARROWHEAD (blue) boundaries on GM12878 (chr 1: 70600000-71530000). The black line represents the predicted probability of each base being a TAD boundary. A zoomed-in portion of the genome shows a yellow-highlighted cluster of bases with high probability of being a boundary, the preciseTAD boundary region (PTBR). (B) Enriched heatmaps comparing the enrichment of CTCF signal around ARROWHEAD TAD boundaries and preciseTAD boundaries, and (C) around Peakachu loop boundaries and preciseTAD for the entire genome. (D) Log 2 genomic distances from boundaries to nearest CTCF sites. (E) Venn diagrams illustrating the levels of conservation between TAD/loop boundaries for GM12878 (red) and K562 (blue) identified by preciseTAD (using ARROWHEAD; top left), ARROWHEAD (top right), preciseTAD (using PEAKACHU; bottom left), and PEAKACHU (bottom right). The median Jaccard Index (IQR) across each chromosome is provided below each venn diagram.](Figures/figure5.png)

@@

@@

**Supplementary Table 2.** TAD/loop boundary data and class imbalance summaries across resolutions for ARROWHEAD and PEAKACHU on the k562 cell line. Class imbalance is calculated as the ratio of total number of unique boundaries to the total number of genomic bins.

**Supplementary Table 3.** A complete list of the genomic annotations used to build the predictor space for all downstream models. The GRCh37/hg19 human genome assembly was used. “Genomic Class” - broad category of genomic features, “Element” - names of genomic features, “Cell Type-Specific Source” - download URL specific to the cell line (not all annotations were provided by the same institutions).

**Supplementary Table 4.** Additional performance metrics when implementing a random forest using
ARROWHEAD ground truth TAD boundaries. Performances are averaged across each holdout chromosome that was reserved for testing.

**Supplementary Table 5.** Additional performance metrics when implementing a random forest using
PEAKACHU ground truth chromatin loop boundaries. Performances are averaged across each holdout chromosome that was reserved for testing.

@@

**Supplementary Figure 2. Distance-type predictors combined with random under-sampling at 5 kb resolution genomic bins yielded optimal TAD boundary region predictive performance for K562 ARROWHEAD TAD boundaries.** Averaged balanced accuracies are compared across resolution, within each predictor-type: (A) Signal, (B) OC, (C) OP, and (D) Distance, and across re-sampling techniques: no re-sampling (None; red), random over-sampling (ROS; green), random under-sampling (RUS; blue), and synthetic minority over-sampling (SMOTE; purple). Error bars indicate 1 standard deviation away from the mean performance across each holdout chromosome used for testing.

**Supplementary Figure 3. Distance-type predictors combined with random under-sampling or SMOTE re-sampling yielded optimal predictive performance PEAKACHU loop boundaries at 10 kb resolution.** Averaged balanced accuracies are compared for A) GM12878 and B) K562  within each predictor-type: Signal, OC, OP, and Distance, and across resampling technique: no re-sampling (None; red), random over-sampling (ROS; green), random under-sampling (RUS; blue), and synthetic minority over-sampling (SMOTE; purple). Error bars indicate 1 standard deviation away from the mean performance across each holdout chromosome used for testing.

**Supplementary Figure 4. Binding of SMC3, RAD21, CTCF, and ZNF143 transcription factors accurately predicts TAD and loop boundaries in K562 cell line.** (A) Barplots comparing performances of TAD (ARROWHEAD) and loop (PEAKACHU) boundary prediction models using histone modifications (HM), cis-regulated elements (BroadHMM), transcription factor binding sites (TFBS), in addition to a model containing all three classes (ALL). (B) Recursive feature elimination (RFE) analysis used to select the optimal number of predictors. Error bars represent 1 standard deviation from the cross-validated accuracy across each holdout chromosome. (C) Clustered heatmap of the predictive importance for the union of the top 8 most predictive chromosome-specific TFBSs. The columns represent the holdout chromosome excluded from the training data. (D) Receiver operating characteristic (ROC) curves and subsequent area under the curves (AUCs) when training and testing on the same cell line versus across cell lines. The curves represent the average sensitivities and specificities across each holdout chromosome.

**Supplementary Figure 5. preciseTAD boundaries are more enriched for known molecular drivers of 3D chromatin.** Enriched heatmaps comparing the enrichment of preciseTAD (green) boundaries with ARROWHEAD called TAD boundaries (blue) around key (A) GM12878-specific and (B) K562-specific transcription factor binding sites known to contribute to the 3D chromatin architecture.

**Supplementary Figure 6. preciseTAD boundaries are more enriched for known molecular drivers of 3D chromatin.** Enriched heatmaps comparing the enrichment of preciseTAD (green) boundaries with PEAKACHU predicted chromatin loop boundaries (blue) around key (A) GM12878-specific and (B) K562-specific transcription factor binding sites known to contribute to the 3D chromatin architecture.

**Supplementary Figure 7. preciseTAD boundaries are spatially closer to known molecular drivers of 3D chromatin.** Boxplots comparing the $log_{2}$ genomic distances from predicted and called boundaries to the nearest (A) GM12878-specific and (B) K562-specific CTCF, RAD21, SMC3, and ZNF143 transcription factor binding sites.

**Supplementary Figure 8. preciseTAD predicted TAD boundaries using ARROWHEAD are nested within preciseTAD predicted chromatin loop boundaries using PEAKACHU.** Venn diagrams illustrating the agreement between boundaries predicted using ARROWHEAD- and Peakachu-trained models. TAD/loop boundaries were flanked by resolution (5 kb for models using ARROWHEAD and 10 kb for PEAKACHU).
